
# Building an R package

Here we present how to efficiently create and build a package using *RStudio* IDE, the `devtools` and `usethis` packages.

More details are provided in the reference material on this subject: the book [*R packages*](https://r-pkgs.org/)[^1] by Hadley Wickham & Jennifer Bryan, freely available online.

[^1]: *R Packages* (2^nd^ Edition) by Hadley Wickham, Jennifer Bryan. O'Reilly, 2023. ISBN: 9781098134945. [https://r-pkgs.org/](https://r-pkgs.org/)


## Initializing a package

A simple way, built into *RStudio*, to **initialize a package** is to follow those steps:

> `r emoji::emoji("point_right")` ***Your turn*** (already)!
>
>  1. Create a new project (top left `File` drop down menu in *RStudio*)
>  
>  2. Choose "*New Directory*"
>  
>  3. Choose "*R package using devtools*" (if it is not available, choose "*R package*", the difference being that with "*R package*" you will have to delete unnecessary files that are automatically created but not useful)
>  
>  4. Give a name to your package, for example `mypkgr`.
  
**NB:** if you are using [Posit Cloud](https://posit.cloud/) and not your local computer, you need to rather execute the following command into the `r fontawesome::fa("r-project")` console: `usethis::create_package("mypkgr")` and accept the potential proposed overwrites.
  
From that, we get the **minimal structure for an** `r fontawesome::fa("r-project")` package, namely:

  - a **DESCRIPTION** file whose `Title`, `Version`, `Authors@R` and `Description` are to be edited (other parts can be edited or even added automatically, see below)
  
  - a **NAMESPACE** file which will be **later** edited automatically (so hands off for now)
  
  - a folder **R/** in which we will add `.R` script files
  
  
`devtools` also adds three ***optional* files**:

  - **.gitignore**, relative to `git`, a version control tool that we will see in detail in the following section on `git` & *GitHub*
  
  - **mypkgr.Rproj** which is a specific file of *RStudio* and allows to define the characteristics and preferences of the project we just created
  
  - **.Rbuildignore** which allows to ignore some files when we build the package down the road (for example, the `mypkgr.Rproj` file should not be included in the package)


## Adding a function: common theme example

We first invite you to code the following function, which we will use throughout the course.

We want to compute the value of the density of a multivariate normal distribution on $\mathbb{R}^p$ at $n$ points. Our function must be applicable for any multivariate normal distribution (i.e. any mean vector $\boldsymbol\mu$ in $\mathbb{R}^p$ and variance-covariance matrix $\boldsymbol\Sigma$ of size $p$) and we wish to compute all the values of the density (eventually in log-scale) evaluated at the $n$ points $\mathbf{x}$ in a single call of the function.

As a reminder, the density function for a multivariate Gaussian distribution is:
$$\displaystyle (2\pi )^{-p/2}\det({\boldsymbol {\Sigma }})^{-1/2}\,\exp \left(-{\frac {1}{2}}(\mathbf {x} -{\boldsymbol {\mu }})^{\mathsf {T}}{\boldsymbol {\Sigma }}^{-1}(\mathbf {x} -{\boldsymbol {\mu }})\right)$$

So you need to create a function `mvnpdf()` in a file named `mvnpdf.R` in the `R/` folder of your package, which :

  - Takes as arguments:
    
      - `x` a matrix, with $n$ columns (the observations) and $p$ rows
      
      - `mean` a vector of means
      
      - `varcovM` a variance-covariance matrix
      
      - `Log` a logical parameter, with default value to `TRUE`.
      
  - Returns a list containing the matrix `x`, and a vector of length $n$ of the multivariate normal distribution density values at those points (eventually in log-scale).

> `r emoji::emoji("point_right")` ***Your turn!***

Here is a function proposal that you can download [here](https://endeavr.borishejblum.science/endeavR_files/mvnpdfRAW.R). `r emoji::emoji("warning")` *WARNING!* if you click too quickly on this link, it will invalidate your participation in the class!

For advice on writing code, see the [R code](https://r-pkgs.org/code.html) section from Wickham & Bryan *R packages* (2023) by Wickham & Bryan.


## Documenting a function

It is important to properly document your code. Every project has at least two developers:

  - yourself
    
  - yourself in 6 months.
  
For the sake of your future self, do yourself a favor and take the time to document your code!

We strongly advise you to use the `roxygen2` package to document your packages. The main advantage being to have the help of a function in the same file as the code defining this function.

> `r emoji::emoji("point_right")` ***Your turn!***  
>
  1. Start by inserting the skeleton of the help with "*Insert Roxygen Skeleton*" located in the "*Code*" menu or the *Magic Wand* sub-menu in the script window.
>  
  2. Complete the documentation by filling in:
>
>      - the title of the function (first line)
>      
>      - the description of what the function does (second paragraph)
>      
>      - if you fill in a third paragraph, this part will go in the "Details" section of the help page
 >     
 >     - the meaning of the parameters (add it after a blank space next to the name of the parameter)
 >     
 >     - the output, after the `@return` tag
 >     
  3. Generate the documentation using "Document" in the "More" menu of the "Build" tab (or alternatively run `devtools::document()` or use `Ctrl+Shift+D`). The effect of this command is multiple:
>
>      - a `man` folder is created and inside it, a `mvnpdf.Rd` file contains the help information about the function `mvnpdf()`
>      
>      - the `NAMESPACE` file is modified
      
**In case of problems OR out of curiosity but only once you are done**, you can have a look at this [proposal](https://endeavr.borishejblum.science/endeavR_files/mvnpdfRox.R).
      
For more details on package documentation and `roxygen2` tags, see the  (comprehensive) [Object documentation](https://r-pkgs.org/man.html) section from Wickham & Bryan *R packages* (2023). Documentation in `r fontawesome::fa("r-project")` uses a specific syntax, `.Rd` (that shares some resemblances with basic $\LaTeX$) (most common commands for text formatting used to be described in previous versions of the `roxygen2` companion package, [see here](https://roxygen2.r-lib.org/articles/formatting.html)). The recommended approach to formating is rather  [enable `markdown` support as described here](https://roxygen2.r-lib.org/articles/rd-formatting.html), which might be more readable in your source file.

Let us finish by mentioning a function from the `usethis` package which initializes a home help page for the whole package:

```{r, eval=FALSE}
usethis::use_package_doc()
```

The generated help page will then be accessible, once the package is installed, with the following command:

```{r, eval=FALSE}
?mypkgr
```


## Interactively test the package

To test the package, you have to load it in `r fontawesome::fa("r-project")` using "Load All" (`Ctrl+Shift+L`) in the "More" menu from the "Build" tab, or alternatively `devtools::load_all()`.

You can then use your package directly in `r fontawesome::fa("r-project")`: look at the function help page with `?mvnpdf` and for example execute the commands given in the example section of this help page.

```{r, eval=FALSE}
?mvndpf
```
 
During code development, you can thus:

  - Add/Modify the `r fontawesome::fa("r-project")` code
  
  - Re-load the package `Ctrl+Shift+L`
  
  - Experiment with it in the console
  
  - And so on...


## Install your package

For the moment, the package exists only in the environment associated with the *RStudio* project we have created. To be able to use it anywhere in `r fontawesome::fa("r-project")` (on your computer), you have to install it (like a CRAN package for example).

To do this, click on "Install" (`Ctrl+Shift+B`) from the "Build" tab, or alternatively you can run `devtools::install()`.

And finally, you can configure *RStudio*'s behavior so that at the time of installation, it simultaneously documents the package: go to the "More" menu from the "Build" tab, then "Configure Build Tools...". Then click on "Configure" next to "Generate documentation with Roxygen", and check the box "Install and Restart" (if it is not already ticked).


## Automatically test your package

To initialize automatic testing capabilities in your package, execute the following command:

```{r, eval=FALSE}
usethis::use_testthat()
```

This command creates a `tests` folder which includes a `testthat.R` file -- not to be modified -- and a `testthat/` folder in which we will add our automated tests. This tool is based on the theory of `unit tests`.

For example, here is the content of a file, containing two tests, that should be named `test-mvnpdf.R` and be put into the `testthat` folder (instead of doing this by hand, you can simply use the helper function `usethis::use_test()` that will create the file at the right place for you):

```{r, eval=FALSE}
test_that("correct result for univariate gaussian", {
  expect_equal(mvnpdf(x=matrix(1.96), Log=FALSE)$y, dnorm(1.96))
  expect_equal(mvnpdf(x=matrix(c(1.96, -0.5), ncol = 2), Log=FALSE)$y,
               dnorm(c(1.96, -0.5)))
})

test_that("correct results for bivariate gaussian", {
  expect_equal(mvnpdf(x=matrix(rep(1.96,2), nrow=2, ncol=1), Log=FALSE)$y,
               mvtnorm::dmvnorm(rep(1.96, 2)))
})
```

To execute these tests, you can click on "Test package" (`Ctrl+Shift+T`) in the "More" menu from the "Build" tab, or run `devtools::test()`.

The advantage of these automatic tests is that they will be run every time you *check* the package (see just below).

A good practice is to add a unit test each time a bug is identified and fixed,
so that we can immediately identify and prevent the same error from happening again in the future.


## Check your package

Running a *check* means ensuring that everything in your package is correct and will work as expected, and that it can be installed properly under various OS. "Passing `R CMD CHECK`" is **mandatory** for successfully uploading your package on CRAN.

To perform `R CMD CHECK` on your package, you can click on "Check" (`Ctrl+Shift+E`) from the "Build" tab, or run `devtools::check()`.

During the *check*, the automated unit tests that we have developed previously are executed. This is the advantage of having written these tests, we don't need to worry about it, but simply react and adjust the code when errors are returned and flagged.



## Appendix 1.1: add an `S3 method` {-}

In most packages, we often have to implement so called `S3 methods` so that, from a result object `res`, one can run `print(res)`, `summary(res)`, `plot(res)`, ...

Here is an example of a `plot()` method that we can add to our package:

```{r, eval=FALSE}
#' Plot of the mvnpdf function
#'
#' @param x an object of class \code{mvnpdf} resulting from a call of the
#' \code{mnvpdf()} function.
#' @param ... graphical parameters passed to \code{plot()} function.
#'
#' @return Nothing is returned, only a plot is given.
#' @export
#'
#' @examples
#' pdfvalues <- mvnpdf(x=matrix(seq(-3, 3, by = 0.1), nrow = 1), Log=FALSE)
#' plot(pdfvalues)
plot.mvnpdf <- function(x, ...) {
  plot(x$x, x$y, type = "l", ...)
}
```

`r emoji::emoji("warning")` *WARNING!* In order for this *S3 method* to do what we want it to do when we apply it to the result of our function `mvnpdf()`, we have to declare that `mvnpdf()` returns a result of class `mvnpdf`.

Test this function, by executing the example.\
*Don't forget to re-install the package ("Install" or `Ctrl+Shift+B`)*.

Look at the contents of the `man` folder and the changes that have been made to `NAMESPACE`.

*Here is a proposed solution*: the [file](https://endeavr.borishejblum.science/endeavR_files/mvnpdf.R) contains the complete code of the `mvnpdf()` function and the associated `plot()` method.


## Appendix 1.2: submit one's package on [CRAN](https://cran.r-project.org/web/packages/available_packages_by_name.html) {-}

Run the two following commands: `devtools::release()`.
For more details, see Wickham & Bryan's recommended pipeline in their book [*R packages*](https://r-pkgs.org/release.html#release-process).
