
# Measuring and comparing execution times

The first step before optimizing a code is to measure its execution time, in order to compare timings between different implementations.

For this section and the following we refer to the book [*Advanced R*](https://adv-r.hadley.nz/) by Hadley Wickham [^3], freely available online.

[^3]: *Advanced R* (2^nd^ Edition) by Hadley Wickham. The R series, CRC press, 2019. ISBN: 9780815384571 [https://adv-r.hadley.nz/](https://adv-r.hadley.nz/)
 
## Measuring execution times with `system.time()`

To measure the execution time of an `r fontawesome::fa("r-project")` command, you can use the native `system.time()` function like this:

```{r, echo=FALSE}
library(mypkgr)
```

```{r}
obs <- matrix(rep(1.96, 2), nrow=2, ncol=1)
system.time(mvnpdf(x=obs, Log=FALSE))
```

The problem that appears in this example is that the execution is so fast that `system.time()` displays `0` (or a very close value) that will be impossible to compare to an hopefully faster implementation. Also, we see that there is some variability when we run the command several times.

Thus if we want to compare our code with the `mvtnorm::dmvnorm()` function, we can't use `system.time()`:

```{r}
obs <- rep(1.96, 2)
system.time(mvtnorm::dmvnorm(obs))
```

We could imagine that we need to increase the complexity of our calculation to be able to compare them, but there is actually a better way: the use of the `microbenchmark` package!



## Compare execution times with `microbenchmark()`

As its name indicates, this package allows to compare execution times even when they are very fast. Moreover, each benchmarked expression will be repeatedly evaluated several times, thus stabilizing its timing estimations.

```{r}
library(microbenchmark)
mb <- microbenchmark(mvtnorm::dmvnorm(rep(1.96, 2)),
                     mvnpdf(x = matrix(rep(1.96,2)), Log = FALSE),
                     times = 1000)
mb
```

```{r, echo=FALSE}
library(ggplot2)
data2plot <- cbind.data.frame("Time" = mb$time/10^6, "Expression" = mb$expr)
levels(data2plot$Expression) <- gsub("mvtnorm::", "", sapply(strsplit(levels(data2plot$Expression), "(", fixed=T), "[", 1))
rangeTime <- c(floor(log10(min(data2plot$Time))):ceiling(log10(max(data2plot$Time))))
brk <- NULL
for(i in rangeTime){
    brk <- c(brk, seq(from = 10^i, to = 10^(i+1), by=10^i))
}
ggplot(data2plot) + geom_violin(aes(x=Expression, y=Time, fill=Expression), alpha=0.8) + 
  scale_fill_manual(guide="none", values=viridis::viridis(6)[1:2]) + 
  scale_y_log10(minor_breaks=brk) + 
  ylab("Execution timings (milli-sec)") +
  xlab("Computed expression") +
  annotation_logticks(sides="l") +
  theme_bw()
```


Both `mvnpdf()` and `dmnvorm()` functions being able to take a matrix as input, we can also compare their speeds in this setting:

```{r}
n <- 100
mb <- microbenchmark(mvtnorm::dmvnorm(matrix(1.96, nrow = n, ncol = 2)),
                     mvnpdf(x=matrix(1.96, nrow = 2, ncol = n), 
                            Log=FALSE),
                     times=100)
mb
```

```{r, echo=FALSE}
library(ggplot2)
data2plot <- cbind.data.frame("Time" = mb$time/10^6, "Expression" = mb$expr)
levels(data2plot$Expression) <- gsub("mvtnorm::", "", sapply(strsplit(levels(data2plot$Expression), "(", fixed=T), "[", 1))
rangeTime <- c(floor(log10(min(data2plot$Time))):ceiling(log10(max(data2plot$Time))))
brk <- NULL
for(i in rangeTime){
    brk <- c(brk, seq(from = 10^i, to = 10^(i+1), by=10^i))
}
ggplot(data2plot) + geom_violin(aes(x=Expression, y=Time, fill=Expression), alpha=0.8)  +
  scale_fill_manual(guide="none", values=viridis::viridis(6)[1:2]) + 
  scale_y_log10(minor_breaks=brk) + 
  ylab("Execution timings (milli-sec)") +
  xlab("Computed expression") +
  annotation_logticks(sides="l") +
  theme_bw()
```


Something happened... And we will find out what exactly is causing this issue in what comes next.
