
# Profiling the code

```{r, echo=FALSE}
library(mypkgr)
library(microbenchmark)
library(ggplot2)
```

*Profiling* is about determining which part of the code takes the most time to compute (and also memory-wise). Once we have found the block of code that takes the longest time to execute, our goal is only to optimize that small part of the code.

To get a *profiling* of the code below, select the lines of code of interest and go to the "*Profile*" menu then "*Profile Selected Lines*". It uses the `profvis()` function of the package `profvis`. 

```{r, eval=FALSE}
n <- 10e4
pdfval <- mvnpdf(x=matrix(1.96, nrow = 2, ncol = n), Log=FALSE)
```

```{r, echo=FALSE}
n <- 10e4
profvis::profvis(mvnpdf(x=matrix(1.96, nrow = 2, ncol = n), Log=FALSE))
```

*OK, OK, we get it!* Concatenating a vector as you go in a loop is really not a good idea in `r fontawesome::fa("r-project")`.


## Comparison with an improved implementation of `mnvpdf()`.

Consider a new version of `mvnpdf()`, called `mvnpdfsmart()`. Download the [file](https://endeavr.borishejblum.science/endeavR_files/mvnpdfsmart.R), and then include it in the package.

Now profile the following command:

```{r, eval=FALSE}
n <- 10e4
pdfval <- mvnpdfsmart(x=matrix(1.96, nrow = 2, ncol = n), Log=FALSE)
```

```{r, echo=FALSE}
n <- 10e4
profvis::profvis(mvnpdfsmart(x=matrix(1.96, nrow = 2, ncol = n), Log=FALSE))
```

We have indeed removed the main computational bottleneck, and we can now learn in a more detailed way what takes time in our function.

To confirm that `mvnpdfsmart()` is indeed much faster than `mvnpdf()` we can make a comparison using `microbenchmark()`:

```{r}
n <- 1000
mb <- microbenchmark(mvnpdf(x = matrix(1.96, nrow = 2, ncol = n), Log = FALSE),
                     mvnpdfsmart(x = matrix(1.96, nrow = 2, ncol = n),
                                 Log = FALSE),
                     times=100L)
mb
```

```{r, echo=FALSE}
data2plot <- cbind.data.frame("Time" = mb$time/10^6, "Expression" = mb$expr)
rangeTime <- c(floor(log10(min(data2plot$Time))):ceiling(log10(max(data2plot$Time))))
brk <- NULL
for(i in rangeTime){
    brk <- c(brk, seq(from = 10^i, to = 10^(i+1), by=10^i))
}
levels(data2plot$Expression) <- gsub("mvtnorm::", "", sapply(strsplit(levels(data2plot$Expression), "(", fixed=T), "[", 1))
ggplot(data2plot) + geom_violin(aes(x=Expression, y=Time, fill=Expression), alpha=0.8)  +
  scale_fill_manual(guide="none", values=viridis::viridis(6)[2:3]) + 
  scale_y_log10(minor_breaks=brk) +
  ylab("Execution timings (milli-sec)") +
  xlab("Computed expression") +
  annotation_logticks(sides="l") +
  theme_bw()
```

We can also check whether `mvnpdfsmart()` becomes competitive with `dmvnorm()`:

```{r}
n <- 1000
mb <- microbenchmark(mvtnorm::dmvnorm(matrix(1.96, nrow = n, ncol = 2)),
                     mvnpdf(x=matrix(1.96, nrow = 2, ncol = n), Log=FALSE),
                     mvnpdfsmart(x=matrix(1.96, nrow = 2, ncol = n), Log=FALSE),
                     times=100L)
mb
```

```{r, echo=FALSE}
library(ggplot2)
data2plot <- cbind.data.frame("Time" = mb$time/10^6, "Expression" = mb$expr)
levels(data2plot$Expression) <- gsub("mvtnorm::", "", sapply(strsplit(levels(data2plot$Expression), "(", fixed=T), "[", 1))
rangeTime <- c(floor(log10(min(data2plot$Time))):ceiling(log10(max(data2plot$Time))))
brk <- NULL
for(i in rangeTime){
    brk <- c(brk, seq(from = 10^i, to = 10^(i+1), by=10^i))
}
ggplot(data2plot) + geom_violin(aes(x=Expression, y=Time, fill=Expression), alpha=0.8)  +
  scale_fill_manual(guide="none", values=viridis::viridis(6)[1:3]) + 
  scale_y_log10(minor_breaks=brk) + 
  ylab("Execution timings (milli-sec)") +
  xlab("Computed expression") +
  annotation_logticks(sides="l") +
  theme_bw()
```

There is still work to be done...



## Comparison with an optimized pure `r fontawesome::fa("r-project")` implementation

After several research, tests, trials and errors, Boris arrived at an [optimized version](https://endeavr.borishejblum.science/endeavR_files/mvnpdfoptim.R) using `r fontawesome::fa("r-project")` capabilities.

Include this `mvnpdfoptim()` function in your package, and then profile it:

```{r}
n <- 10e4
profvis::profvis(mvnpdfoptim(x=matrix(1.96, nrow = 2, ncol = n), Log=FALSE))
```

And the `microbenchmark()` that goes with it:

```{r}
n <- 1000
mb <- microbenchmark(mvtnorm::dmvnorm(matrix(1.96, nrow = n, ncol = 2)),
                     mvnpdf(x=matrix(1.96, nrow = 2, ncol = n), Log=FALSE),
                     mvnpdfsmart(x=matrix(1.96, nrow = 2, ncol = n), Log=FALSE),
                     mvnpdfoptim(x=matrix(1.96, nrow = 2, ncol = n), Log=FALSE),
                     times=100L)
mb
```

```{r, echo=FALSE}
library(ggplot2)
data2plot <- cbind.data.frame("Time" = mb$time/10^6, "Expression" = mb$expr)
levels(data2plot$Expression) <- gsub("mvtnorm::", "", sapply(strsplit(levels(data2plot$Expression), "(", fixed=T), "[", 1))
rangeTime <- c(floor(log10(min(data2plot$Time))):ceiling(log10(max(data2plot$Time))))
brk <- NULL
for(i in rangeTime){
    brk <- c(brk, seq(from = 10^i, to = 10^(i+1), by=10^i))
}
ggplot(data2plot) + geom_violin(aes(x=Expression, y=Time, fill=Expression), alpha=0.8)  +
  scale_fill_manual(guide="none", values=viridis::viridis(6)[1:4]) +
  scale_y_log10(minor_breaks=brk) + 
  ylab("Execution timings (milli-sec)") +
  xlab("Computed expression") +
  annotation_logticks(sides="l") +
  theme_bw()
```

Finally,  we can profile the `dmvnorm()` function:

```{r, eval=TRUE}
n <- 10e6
profvis::profvis(mvtnorm::dmvnorm(matrix(1.96, nrow = n, ncol = 2)))
profvis::profvis(mypkgr::my_dmvnorm(matrix(1.96, nrow = n, ncol = 2)))
```

You can download the `my_dmvnorm()` function [here](https://github.com/borishejblum/mypkgr/blob/1df7c6ba37df4ec7406167953beea917757d0481/R/my_dmvnorm.R) and include it in your package, in order to have the source code available in the profiling result.
